Staged Abstract Interpreters

### 2019-08-15 paper/10_intro.tex
revision
Proofreading intro

In recent years, we observe a rich progress on the methodologies for
constructing abstract interpreters from systematic principles, instead
of ad-hoc engineering. A notable one is the Abstracting Abstract
Machine (AAM) methodology \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}. It uncovers an approach to derive sound
abstract interpreters from their concrete counterparts, for example,
abstract machines, where the soundness can be easily established by
examining the transformation of semantic artifacts. For example, the
CEK machine \cite{DBLP:conf/popl/FelleisenF87} for concrete execution
can be readily refactored to an effective $0$-CFA control-flow
analysis \cite{Shivers:1988:CFA:53990.54007,
  Midtgaard:2012:CAF:2187671.2187672}: first we tweak the environment
dereference as a nondeterministic choice, such that the environment
may contain mutiple possible values for a varaiable, then allocating
continuations in the environment to exploit return-flow information,
and also constrain the addresses space to be finite.  This systematic
abstraction approach can be tailored to different language features
(such as states, first-class controls, exceptions and concurrency) and
sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
also has been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

Based on the idea of abstracting abstract machines, more
pragmatically, several implementation strategies utilizing purely
functional programming to build abstract interpreters were
emerged. Such techniques include monads, monad transformers
\cite{DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979},
arrows \cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Githubsemantic} and etc.  The pure approaches provide certain
benefits. The abstract interpretation artifacts can be built
compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proved with
less efforts, either by mechanized \cite{Darais:2016:CGC:2951913.2951934}
or paper-based proof \cite{Keidel:2018:CSP:3243631.3236767}. Also,
referential transparent and equational reasoning allow programmers to
reason the correctness of their implementations more confidently.

However, besides the intrinsic complexity of static analysis, there
are additional abstraction penalties with these high-level
implementation approaches.  First, similar to concrete interpreters,
the abstract interpreter analyzes the program by traversing the
abstract syntax tree, which poses an interpretive overhead, such as
the pattern matching on the ASTs and recursive calls on the sub
expressions. Those kind of overhead can be negligible if the abstract
interpreter only runs on the program for one time, but also can be
accumulated significantly if it runs repeatedly, for example, on
libraries.  Second, the abstract interpreter written in pure languages
usually extensively uses effect systems to implement the semantics of
abstract interpretation.  For example, the abstract interpreter that
returns all possible runtime values is a form of nondeterminism, where
nondeterminism monads can help.  Although such pure approaches have
its own merits and elegance, compared with imperative stateful
implementations, they are significantly slower.

In this paper, we propose an abstraction-without-regret approach to
eliminate those performance penalties for abstract interpreters,
meanwhile still keeping the benefits come from purely functional
programming.  In short, our approach applies ideas and techniques from
program specialization and embedded DSLs to abstract interpreters. 1)
Using multi-stage programming , we can specialize the abstract
interpreter with respect to an input program and then generate
efficient low-level code that does the actual analysis.  The result of
specialization is reusable, and the effect layers have been eliminated
in the generated code. 2) Inspired by the tagless-final interpreters,
we define a generic interpreter that abstracts over binding-time and
different semantics, which allows user to implement different
semantics modularly, including the staged abstract interpretation
semantics.  Together with the multi-stage programming and type-level
annotations of binding-times, we can derive the staged abstract
interpreters without intrusive changes to the corresponding unstaged
one, thus the soundness is untouched.  In this sense, our approach is
no regret of both performance and engineering effort. We elaborate
these two main ideas in detail as follow.

@@

\paragraph{Applications and Evaluations}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, a implementation technique for
control-flow analyses, and show that by utilizing type-based stage
annotations we can achieve the same optimization, and meanwhile,
the analyzer does not need to change, thereby requires much less
engineering efforts.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis and abstract garbage collection.
3) We notice that staging an abstract interpreter enables modularly
compiling an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore the approach provides a modular way to create optimized
analysis code by mechanized reusing a whole-program analyzer.
4) We also empirically evaluate the performance improved by staging,
showing an order of magnitude speed-up on flow-analysis \todo{explicit
  note the number}.

>>>>

In recent years, we observe rich progress on the methodologies for
constructing abstract interpreters from systematic principles rather than
ad-hoc engineering. Of particular note is the Abstracting Abstract
Machine (AAM) methodology \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}. This methodology uncovers an approach to derive sound
abstract interpreters from their concrete counterparts, for example,
abstract machines, where the soundness can be easily established by
examining the transformation of semantic artifacts. For example, the
CEK machine \cite{DBLP:conf/popl/FelleisenF87} for concrete execution
can be readily refactored to an effective $0$-CFA control-flow
analysis \cite{Shivers:1988:CFA:53990.54007,
  Midtgaard:2012:CAF:2187671.2187672}: first, tweak the environment
dereference as a nondeterministic choice, such that the environment
may contain mutiple possible values for a varaiable, then allocate
continuations in the environment to exploit return-flow information,
and also constrain the addresses space to be finite.  This systematic
abstraction approach can be tailored to different language features
(such as states, first-class controls, exceptions, and concurrency) and
sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
has also been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

Based on the idea of abstracting abstract machines, more
pragmatically, several implementation strategies utilizing purely
functional programming to build abstract interpreters have
emerged. Such techniques include monads, monad transformers
\cite{DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979},
arrows \cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Githubsemantic}, etc.  The pure approaches provide certain
benefits, including that the abstract interpretation artifacts can be built
compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proven with
less effort, either by mechanized \cite{Darais:2016:CGC:2951913.2951934}
or paper-based proofs \cite{Keidel:2018:CSP:3243631.3236767}. Also,
referential transparent and equational reasoning allow programmers to
reason the correctness of their implementations more confidently.

However, besides the intrinsic complexity of static analysis, there
are additional abstraction penalties with these high-level
implementation approaches.  First, similar to concrete interpreters,
the abstract interpreter analyzes the program by traversing the
abstract syntax tree (AST), which poses an interpretive overhead, such as
the pattern matching on the ASTs and recursive calls on the subexpressions.
Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can also
accumulate significantly if run repeatedly (for example, on
libraries).  Second, the abstract interpreter written in pure languages
usually extensively uses effect systems to implement the semantics of
abstract interpretation.  For example, the abstract interpreter that
returns all possible runtime values is a form of nondeterminism, where
nondeterminism monads can help.  Although such pure approaches have
their own merits and elegance, compared with imperative stateful
implementations, they are significantly slower.

In this paper, we propose an abstraction-without-regret approach to
eliminating performance penalties for abstract interpreters,
while still keeping the benefits inherent in purely functional
programming.  In short, our approach applies ideas and techniques from
program specialization and embedded DSLs to abstract interpreters. 1)
Using multi-stage programming , we can specialize the abstract
interpreter with respect to an input program and then generate
efficient low-level code that does the actual analysis.  The result of
specialization is reusable, and the effect layers have been eliminated
in the generated code. 2) Inspired by the tagless-final interpreters,
we define a generic interpreter that abstracts over binding-time and
different semantics, which allows user to implement different
semantics modularly, including the staged abstract interpretation
semantics.  Together with the multi-stage programming and type-level
annotations of binding-times, we can derive the staged abstract
interpreters without intrusive changes to the corresponding unstaged
one, thus the soundness is untouched.  In this sense, our approach is
no regret of both performance and engineering effort. We elaborate
these two main ideas in detail as follow.

@@

\paragraph{Applications and Evaluations}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, a implementation technique for
control-flow analyses, and show that by utilizing type-based stage
annotations we can achieve the same optimization, and meanwhile,
the analyzer does not need to change, thereby requires much less
engineering efforts.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis and abstract garbage collection.
3) We notice that staging an abstract interpreter enables modularly
compiling an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore the approach provides a modular way to create optimized
analysis code by mechanized reusing a whole-program analyzer.
4) We also empirically evaluate the performance improved by staging,
showing an order of magnitude speed-up on flow-analysis.
